Place of death in NRS COVID-19 deaths: data processing for analysis, 2023 data

### 2023-04-20 covid_nrs_place_of_death_data_2023.Rmd
made changes to accommodate the disaggregatd data by age,sex,LA/HB ne longer being available alongside place of death added figure 2 for the paper

@@

```{r}
## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
## Note: as of 16 Feb 2023, this returns a strange error
# TODO: if error persists, load data from the main report file instead?
ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")

@@

```{r}
weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

@@

```{r}
weekly_covid_contributory_vs_underlying_cause <-
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  filter(
    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
  ) %>%
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  clean_names() %>%
  filter(age!="all") %>%
  filter(sex!="all") %>%
  filter(place_of_death != "All") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x,
      x %>% 
        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(place_of_death = "All")
    )
  })

## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure
stopifnot(
  (weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
  filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
  nrow) == 0
)

@@

```{r}
## constants for use in graphs & annotations
weeks_available_2023 <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
most_recent_date_available_2023 <- max(merged_deaths_age$date_w_c)

earliest_date <- min(merged_deaths_age$date_w_c)
current_year <- isoyear(most_recent_date_available_2023)
```

>>>>

TODO: now that the ad-hoc files are no longer available, some of the data will need to be compiled from the tabs in the weekly data.

@@

Note: this contains data split by sex and age, but only for ALL places of death; individual PoD data are available for ALL sex and ALL age only

```{r}
## a separate raw file makes it easier to debug, and also reduces the number of requests I make while debugging!
ods_raw <- ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc")

@@

Note: the last sex & age split of the data provided by NRS was Week 9 of 2023, though the data can still be split using the 'overall' dataset from Open Data Scotland above!

```{r}
# weekly_deaths_sex_age <-
#   weekly_deaths_overall %>%
#   filter(age!="all" & place_of_death != "All")

# TODO: delete the below line, used for testing
# files_weekly_deaths$main_report <- "./downloaded_data/2023-03-22_weekly-deaths-23.xlsx"

sheet_covid_deaths_sex_and_age_data <- "1"
sheet_all_cause_deaths_sex_and_age_data <- "4"

cell_ranges_for_sex_and_age_data <- list(
  "All" = "A6:K110",
  "Female" = "A113:K217",
  "Male" = "A220:K324"
)

# TODO: this still doesn't split data by PoD...
weekly_deaths_sex_age <-
  full_join(
    x =
      map_dfr(
        .x = cell_ranges_for_sex_and_age_data,
        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_covid_deaths_sex_and_age_data, range = .x), 
        .id = "sex"
      ) %>%
      pivot_longer(
        cols = `All ages`:`85+`,
        names_to = "age",
        values_to = "deaths_covid_related"
      ) %>%
      janitor::clean_names(),
    y =
      map_dfr(
        .x = cell_ranges_for_sex_and_age_data,
        .f = ~readxl::read_excel(files_weekly_deaths$main_report, sheet = sheet_all_cause_deaths_sex_and_age_data, range = .x), 
        .id = "sex"
      ) %>%
      pivot_longer(
        cols = `All ages`:`85+`,
        names_to = "age",
        values_to = "deaths_all_causes"
      ) %>%
      janitor::clean_names(),
    by = c("sex", "registration_year", "week_number", "week_beginning", "age")
  ) %>%
  rename(year = registration_year, date_w_c = week_beginning) %>%
  mutate(
    year = as.integer(year),
    date_w_c = as.Date(date_w_c, origin = "1899-12-30"),  # origin to account for Excel's integer date format
    week_number_run_over = compute_week_run_over(week_number, year, start_year = 2020),  # invent a "run-over" week number - weeks 54+ are in 2021, weeks 106+ are in 2022
    deaths_non_covid = deaths_all_causes - deaths_covid_related  # compute non-covid from all-cause minus covid
  ) %>%
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

# TODO: this is obsolete now, remove!
weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  recode_cause_of_death() %>%
  recalculate_number_of_deaths_after_recoding_cause_of_death %>%  # recompute the sum of deaths
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()
```

@@

```{r}
# TODO: need to compute that from another source for the time being
weekly_covid_contributory_vs_underlying_cause <-
  readxl::read_excel(files_weekly_deaths$sex_age, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  parse_year_and_week_number_from_ad_hoc_data %>%
  filter(
    cause_of_death %in% c("COVID-19 contributory factor", "COVID-19 underlying cause")
  ) %>%
  tidyr::complete(  # complete implicit missing entries and make them 0
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  clean_names() %>%
  filter(age!="all") %>%
  filter(sex!="all") %>%
  filter(place_of_death != "All") %>%
  group_by(week_number, week_number_run_over, year, date_w_c, place_of_death) %>%  # sum over age & sex
  summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
  ({function(x)
    bind_rows(
      x,
      x %>% 
        group_by(week_number, week_number_run_over, year, date_w_c) %>%  # sum across place of death
        summarise(across(starts_with("covid_"), ~sum(.x)), .groups = "drop") %>%
        mutate(place_of_death = "All")
    )
  })

# TODO: this is currently commented out as it doesn't compute correctly
## check that the figures actually make sense! i.e. the two figures add up to the total covid-related figure

# stopifnot(
#   (weekly_deaths_overall %>%
#   filter(sex=="all" & age=="all") %>%
#   inner_join(weekly_covid_contributory_vs_underlying_cause) %>%
#   filter(deaths_covid_related != covid_19_contributory_factor + covid_19_underlying_cause) %>%
#   nrow) == 0
# )

@@

There is as of April 2023 a split in the availability of data:

- Data are being continually updated for the all age all sex all HB/LA dataset
- Data split by age, sex, HB or LA stopped being updated (technically they are still available but not without place of death data)

Therefore there are two separate most recent available data, I'll call them 'basic' and 'disaggregated'/

```{r}
## constants for use in graphs & annotations
weeks_available_2023 <- max(merged_deaths_overall$week_number[merged_deaths_overall$year==2023])
most_recent_date_available_2023 <- max(merged_deaths_overall$date_w_c)

earliest_date <- min(merged_deaths_overall$date_w_c)
current_year <- isoyear(most_recent_date_available_2023)

weeks_available_2023_disaggregated <- max(merged_deaths_age$week_number[merged_deaths_age$year==2023])
most_recent_date_available_2023_disaggregated <- max(merged_deaths_age$date_w_c)
```
